Lossy compression of DNA sequencing quality data

### 2015-04-22 manuscript/q-compression.tex
Added beginning supplemental spreadsheet.

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
sets, but the additional short-jump library results are available in
the supplementary (TODO:TABLE). Storing the uncompressed quality
values requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black asterisk in
the figure. The lossless compression of each data set using \textsc{bz}ip2
ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored
asterisks on the abscissa. The compression methods tend to cluster
together across the different data sets. Across all data sets, the
0-degree polynomial regression, profile encodings, and \textsc{q}ual\textsc{c}omp have
the lowest bits/base-call.

@@

It is important to highlight that the even though a compression method
may result in the same number of filtered base-pairs as the
uncompressed sequences, it does not mean the \emph{same} base-pairs
were filtered. The 1st-degree and 5th-degree polynomial regression of
the \textit{Rhodobacter sphaeroides} fragment library filters nearly
as many bases as each other. However, if we examine the specific
reads filtered, the 5th-degree polynomial regression discards
approximately two thirds less reads than the 1st-degree polynomial
regression that are kept by the uncompressed reads (4,640 and
12,066 reads, respectively) (TODO: move to Supplemental Table
\ref{tab:quality_control}).

@@

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
values. All compression methods other than 0-degree polynomial
regression produce an N50 ranging from 3.17--3.22 Mbps (see
Supplemental Table TODO). Despite the similar contiguity statistics,
the different compression methods vary noticeably in the amount of
\textsc{snp}s. The order of polynomial has an inverse relationship
with the amount of \textsc{snp}s detected. The 2-bin and profile
methods detected the least amount of \textsc{snp}s compared to the
reference genome, outperforming the assembly using the original
quality values. A more in-depth evaluation is needed to determine
whether these compression methods are missing actual \textsc{snp}s.

It is important to highlight that using uncompressed reads does not
produce the best assembly in terms of any of the metrics. The
uncompressed reads scores worse than the top overall assembly
(256-profile encoding) in terms of assembled bases, missing reference
bases, N50, \textsc{snp}s, indels $>$5bp, and relocations. The
assembly using uncompressed reads has an error rate of roughly 8.75
errors per 100 kb of assembled sequence, while the 256-profile
encoding has an error rate of 8.02 errors per 100 kb (see Supplemental
table TODO).

>>>>

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides} 2.4.1, \textit{Homo sapiens}
chromosome 14, \textit{Escherichia coli} str. K-12 MG1655, and
\textit{Mus musculus} data sets (Figure \ref{fig:mse_vs_bpbp}). We
only include the fragment libraries for the \textit{Rhodobacter
  sphaeroides} 2.4.1, and \textit{Homo sapiens} chromosome 14 data
sets, but the additional short-jump library results are available in
the Supplemental Table 1. Storing the uncompressed quality
values requires 1 byte per base-call because they are stored in
\textsc{ascii} format and is denoted by the dotted black asterisk in
the figure. The lossless compression of each data set using \textsc{bz}ip2
ranges from 2.19 - 3.10 bits/base-call and is denoted by the colored
asterisks on the abscissa. The compression methods tend to cluster
together across the different data sets. Across all data sets, the
0-degree polynomial regression, profile encodings, and \textsc{q}ual\textsc{c}omp have
the lowest bits/base-call.

@@

It is important to highlight that the even though a compression method
may result in the same number of filtered base-pairs as the
uncompressed sequences, it does not mean the \emph{same} base-pairs
were filtered. The 1st-degree and 5th-degree polynomial regression of
the \textit{Rhodobacter sphaeroides} fragment library filters nearly
as many bases as each other. However, if we examine the specific
reads filtered, the 5th-degree polynomial regression discards
approximately two thirds less reads than the 1st-degree polynomial
regression that are kept by the uncompressed reads (4,640 and
12,066 reads, respectively) (Supplemental Table 2).

@@

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
values. All compression methods other than 0-degree polynomial
regression produce an N50 ranging from 3.17--3.22 Mbps (see
Supplemental Table 3). Despite the similar contiguity statistics,
the different compression methods vary noticeably in the amount of
\textsc{snp}s. The order of polynomial has an inverse relationship
with the amount of \textsc{snp}s detected. The 2-bin and profile
methods detected the least amount of \textsc{snp}s compared to the
reference genome, outperforming the assembly using the original
quality values. A more in-depth evaluation is needed to determine
whether these compression methods are missing actual \textsc{snp}s.

It is important to highlight that using uncompressed reads does not
produce the best assembly in terms of any of the metrics. The
uncompressed reads scores worse than the top overall assembly
(256-profile encoding) in terms of assembled bases, missing reference
bases, N50, \textsc{snp}s, indels $>$5bp, and relocations. The
assembly using uncompressed reads has an error rate of roughly 8.75
errors per 100 kb of assembled sequence, while the 256-profile
encoding has an error rate of 8.02 errors per 100 kbp (Supplemental
Table 3).
